---
title: "Climate, Crises, and Contenment: A Global Well-Being Analysis"
author: "Group 15"
date: "November 12th, 2024"
output: pdf_document
---

List your group members, including their student numbers, here:

-   Alexa Melnechuk (169086699)
-   Joey Kielt (169087303)
-   Sanjana Mahabir (169080789)
-   Uma Sisira Akella (200300920)
-   Shalin Panjwani (169073107)

You **must** be in a group in MyLS in order to see the DropBox used for submission. Even if you're alone, you must join a group by yourself.

You **must** be in a group with people from the same section as you. MyLS does not allow for groups including students from both Data100A and Data100B.

```{r setup, include=FALSE}
# echo = FALSE will set the Rmd to *not* show the R code. Don't change this.
# You may change the default figure width and figure height as you please.
knitr::opts_chunk$set(echo = FALSE, message = FALSE, fig.width = 6)

# Put any libraries that you need to load here.
# DO NOT PUT "install.packages()" IN AN RMD FILE!!!
library(tidyverse)
library(arrow)
library(openxlsx)
library(ggridges)
library(ggplot2)
library(viridis)
library(dplyr)
library(lubridate)
theme_set(theme_bw())
```

# Abstract

This report focuses on how global challenges such as climate change awareness, natural disasters and the COVID-19 pandemic’s impact on national well-being. We explore how these factors correlate and potentially influence each other by analyzing data on happiness scores, climate awareness, COVID-19 cases, sea ice extent, and cyclone strength. The research aims to uncover patterns in how environmental consciousness, extreme weather events, and global health crises affect psychological resilience and overall happiness in various cultural and geographical contexts. The data used was cleaned and then further manipulated to emphasize the correlation between the different data sets. We started off by finding out if climate awareness correlated with happiness across countries. Then we explored whether there was a link between cyclone strength and sea ice extent. Finally, we explored whether higher COVID-19 cases in 2020 impact national happiness levels, with variations by climate awareness. Through this analysis, we came to the conclusion that global challenges such as climate change, natural disasters, and pandemics do indeed affect national well-being, albeit with different effects across regions.

# Introduction

Introduction: 

In this report, we will mainly be exploring the relationship between climate change, how aware people are of it, and well-being. Our key research questions are:

1.  How does climate change awareness correlate with happiness in different parts of the world?

2.  Is there a connection between Arctic sea ice loss, hurricane intensification, and climate awareness?

3.  Did countries with higher COVID-19 case rates in 2020 report lower happiness levels? Does this relationship vary based on climate awareness?

The data we analyze includes happiness scores, climate awareness percentages, COVID-19 case counts, Arctic sea ice extent and cyclone intensity. We will be using exploratory data analysis techniques including scatterplots, ridgeline plots, and grouped boxplots. Using these we can identify patterns and relationships between various variables. 

By the end of this report, we will have gained various insights into how climate awareness, hurricane intensity, and the other factors we explored could relate to happiness and well-being across regions. An understanding of this data can help in guiding future efforts to provide better support to communities facing or impacted by crises.

# Data Description

## \<\<Data Set 1\>\>

```{r load_data1}
cyclones_data <- read_parquet("./cyclones_data.parquet")
cyclones_summary <- cyclones_data |>
  mutate(year = year(date)) |>
  group_by(year, Basin) |>
  summarize(avg_max_wind = mean(max_wind, na.rm = TRUE))

```

The data come from \<<place>\> and describe \<<more specific description of the data>\>.

In order to clean the data, we \<\<steps to clean the data, concise but precise enough that a reader could follow your steps without seeing your code\>\>.

## \<\<Data Set 2\>\>

```{r load_data2}
sea_ice_extent <- read_parquet("./ice_extent_yearly.parquet")
ice_summary <- sea_ice_extent |>
  filter(region %in% c("Arctic", "Antarctic")) |>
  group_by(year, region) |>
  summarize(avg_extent = mean(value, na.rm = TRUE))

```

The data come from \<<place>\> and detail \<<more specific description of the data>\>.

In order to clean the data, we \<\<steps to clean the data, concise but precise enough that a reader could follow your steps without seeing your code\>\>

## \<\<Data Set 3\>\>

```{r load_data3}
climate_awareness <- read_parquet("./climate_awareness.parquet") |>
  mutate(
    aware_score = aware_alot + 0.75*aware_moderate + 0.5*aware_alittle + 0.25*aware_no
    )
climate_happening <- read_parquet("./climate_happening.parquet")

```

The data come from \<<place>\> and detail \<<more specific description of the data>\>.

In order to clean the data, we \<\<steps to clean the data, concise but precise enough that a reader could follow your steps without seeing your code\>\>

## \<\<Data Set 4\>\>

```{r load_data4}
covid_2020 <- read_parquet("./covid_2020.parquet")
```

The data come from \<<place>\> and detail \<<more specific description of the data>\>.

In order to clean the data, we \<\<steps to clean the data, concise but precise enough that a reader could follow your steps without seeing your code\>\>

## \<\<Data Set 5\>\>

```{r load_data5}
happiness_data <- read_parquet("./happiness.parquet")
```

The data come from \<<place>\> and detail \<<more specific description of the data>\>.

In order to clean the data, we \<\<steps to clean the data, concise but precise enough that a reader could follow your steps without seeing your code\>\>

## Combining the Data

Explain how any combinations of data were performed. Explain what kind of join was needed, whether columns had to be modified (for example, matching "country" names.)

```{r}

# 1. Does climate awareness correlate with national happiness levels across continents?

#Combining data
awareness_happiness <- climate_awareness |>
  select(country, aware_alot) |>
  merge(happiness_data |> select(country, life_ladder), by = "country") |>
  merge(covid_2020 |> select(country, continent), by = "country") |>
  filter(!is.na(continent), !is.na(aware_alot), !is.na(life_ladder))

# Create scatter plot
ggplot(awareness_happiness, aes(x = aware_alot, y = life_ladder, color = continent)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Climate Awareness vs. Happiness by Continent",
       x = "Climate Awareness Level",
       y = "Happiness Score",
       color = "Continent") +
  theme_minimal()



# Data Set 1: Does climate awareness correlate with national happiness levels across continents?
# The data from above consists of climate awareness, happiness and covid-19 cases for continent grouping. This data was used to determine whether higher climate awareness correlates with higher or lower happiness scores by country in context with continental grouping to identify if there are regional patterns. Through examining whether countries with higher climate change awareness report higher or lower happiness levels, it helps to develop an understanding of whether climate change implications influence mental well being. By using regional comparisons, it can reveal if the psychological impact of awareness varies culturally or geographically, showcasing if it affects individuals' well being differently across the continents. 
Cleaning process described: First we began by loading the NOAA data files for cyclones in the Atlantic and North Pacific basins. Where then the columns are set up to display the wind speed, pressure, and location coordinates. When dealing with the missing values in the data set "-999" and "-99" were used as placeholders. The code then replaces these values with NA allowing for easier filtering. When structuring the cyclone data set columns like BasinNumberYear, Name, and Entries are created to easily identify the rows. The columns hold data for each cyclone, such as name and year. The fill function is used to carry values down in these columns, ensuring each cyclone entry has complete data. Once both the Atlantic and North Pacific data sets are created, they are combined into a single data frame for us to easily analyze. When cleaning the coordinate data latitude and longitude are initially in text format. Using the function, convert_latlon,we were able to convert it into numeric values. Where the north and east values are positive and south and west values are negative. After completing the cleaning process, the dataset is saved as a parquet file named "cyclones_data.parquet". This format is efficient for storage and can be easily used for further analysis.




# 2. Is there a link between sea ice extent changes and hurricane strength?

# Combining data
combined_data <- merge(cyclones_summary, ice_summary, by = "year")

# Create the ridgeline plot with facets
ggplot(combined_data, aes(x = avg_max_wind, y = region, fill = after_stat(x))) +
  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
  facet_wrap(~ Basin, scales = "free", ncol = 2) +
  scale_fill_viridis_c(name = "Avg Max Wind Speed (kt)", option = "C") +
  labs(title = "Sea Ice Extent vs Hurricane Strength (1979-2023)",
       subtitle = "Ridgeline plot of average maximum wind speed by basin and ice region",
       x = "Average Maximum Wind Speed (kt)",
       y = "Sea Ice Region") +
  theme_ridges() +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))




Data Set 2: Is there a link between sea ice extent changes and hurricane strength?
 The data consists of cyclones for hurricane strength and ice extent in terms of Arctic and Antarctic. This data was used to determine if years with intense hurricanes corresponded to years with lower ice extent – looking at the environmental interdependencies. By analyzing extreme whether events like hurricanes and changing ice extents which are indicators of climate change through looking at the direct impacts to the environment and the indirect effect to national well being by increasing economic and social vulnerabilities. By examining if severe hurricanes correlate with reduced sea ice, it hints at broader environmental implications of climate change – an ecosystem under strain. This ecological strain can translate into social challenges like affecting public health, safety and well being overtime, primarily hurting communities where these disasters are likely to happen. 
Cleaning process described: We loaded the sea ice data from an Excel file containing Northern Hemisphere (NH) and Southern Hemisphere (SH) ice extent data from 1978 to 2023. The code reads the NH sheet, selecting specific columns and renaming them to tidy the data. The data is then tidied further using pivot_longer() to create a single year column, allowing each row to contain month, day, year, and ice_extent. The same cleaning process is also used for the SH sheet. The NH and SH datasets are then combined into one, with a new date column created from year, month, and day, all sorted by region and date. The code generates a line plot of daily sea ice extent by region to visualize annual trends, then calculates yearly minimum and maximum ice extents for each region. These min and max values are organized in a table and saved as "ice_extent_yearly.parquet". 



# 3. Did higher COVID-19 cases in 2020 impact national happiness levels, with variations by climate awareness?

# Combining data
covid_happiness_awareness <- merge(
  merge(
    covid_2020 |> select(country, continent, total_cases),
    happiness_data |> select(country, life_ladder),
    by = "country"
  ),
  climate_awareness |> select(country, aware_alot),
  by = "country"
) |>
  mutate(
    covid_impact = ifelse(total_cases > median(total_cases, na.rm = TRUE), "High", "Low"),
    climate_awareness = ifelse(aware_alot > median(aware_alot, na.rm = TRUE), "High", "Low")
  )

# Create grouped bar plot
ggplot(covid_happiness_awareness, aes(x = covid_impact, y = life_ladder, fill = climate_awareness)) +
  geom_bar(stat = "summary", fun = "mean", position = "dodge") +
  facet_wrap(~continent) +
  labs(title = "COVID-19 Impact on Happiness by Climate Awareness and Continent",
       x = "COVID-19 Impact",
       y = "Average Happiness Score",
       fill = "Climate Awareness") +
  theme_minimal()


```
Data Set 3: Did higher COVID-19 cases in 2020 impact national happiness levels, with variation with climate awareness?
The data contains the COVID-19 cases, happiness and climate awareness. By checking if countries with higher COVID-19 case counts show significantly different happiness scores, and seeing if this effect differs for countries with higher climate awareness.The COVID-19 pandemic has had significant impacts on global well-being, disrupting economies, health systems, and daily life. By analyzing how happiness levels differ in countries with varying COVID-19 impacts, this question assesses the social resilience and mental well-being of populations facing a health crisis. Factoring in climate awareness further explores whether countries with higher environmental consciousness also experience distinct emotional responses to the pandemic, potentially indicating if global awareness shapes mental resilience during crises.
Describe the cleaning process. COVID-19 data from Our World In Data to analyze total cases by country for 2020, combining it with data on climate awareness for comparison. We began by loading the COVID data, selecting specific columns: date, iso_code, location, and continent, and then filtering it to only have 2020 cases. To compare COVID cases with climate awareness, the code aligns country names in both datasets. First, it replaces periods in country names (e.g., “U.S.A.” becomes “USA”) and then corrects specific mismatches (e.g., “Bolivia (Plurinational State of)” is shortened to “Bolivia” and “Hong Kong” becomes “Hong Kong SAR, China”). Once names are standardized, the code joins the COVID data with World Bank population data and calculates each country’s COVID cases per capita by dividing total cases by population.




# Exploratory Data Analysis

To achieve our goals, we explored the data by...

We explored many aspects of the data, but will demonstrate three. These are \<\<insight 1\>\>, \<\<insight 2\>\>, and \<<insight3>\>

The first aspect that we found interesting is shown in \@ref(fig:insight1). The insight should be specific to the data shown, not a general statement beyond the data (leave that for the conclusion).

```{r insight1, fig.cap="This is a figure caption that you will need to change in order to get good marks in the visualization rubric items."}
# This is an example of how you can control figures and captions in
# an R chunk. Note that you can reference figures using:
# \@ref(fig:insight1), where "insight1" is the label of this code
# chunk (the first bit of text after the "r" in "```{r label, options...}")
```

This insight is supported by the summary statistics in table \@ref(tab:summary_stats)

```{r summary_stats}
# Calculate the relevant summary statistics here.
# Note that the "kable" function in the "knitr" package
# is convenient for making nice tables. Other packages can
# do much fancier things with tables, but keep in mind that
# the insights should be the star, not the formatting.
```

The next insight that we found is shown in \@ref(fig:insight2).

```{r insight2, fig.height=4, fig.width=6, fig.cap="This is a figure caption that you will need to change in order to get good marks in the visualization rubric items."}
# This figure will have a height of 4 and a width of 6.
# Feel free to change this, and to apply different sizes
# to the other figures you create.
```

Finally, \@ref(fig:insight3) shows ...

```{r insight3, fig.height=4, fig.width=6, fig.cap="This is a figure caption that you will need to change in order to get good marks in the visualization rubric items."}
```

# Conclusion and Future Work

Overall, we found \<<general ideas>\>.

A second paragraph about our findings.

The next steps in this analysis are...

The limitations of this analysis are as follows. (Do not simply list potential issues with sampling, but relate them to your analysis and how they affect your conclusions. An honest and complete acknowledgement of the limitations makes the analysis more trustworthy.)

# References

I am not strict about MLA or APA style or anything like that. For this report, I would much rather have your citations be easy to match to your insights.

The easiest way is to use Rmd's [footnote](https://bookdown.org/yihui/rmarkdown/markdown-syntax.html#inline-formatting) syntax. This will put a number beside the word where the footnote appears, and the full text of the footnote at the bottom of the page (pdf) or end of the document (html). The syntax is:[^1], where I suggest that you put in something like this[^2] to make references for this assignment.

[^1]: See the source view to see this footnote

[^2]: The relevance to the insight is ... . From \<<name of source and name of article>\>, published on \<<date>\>, url: \<<link to page>\>

Alternatively, you could make a list of citations with their main arguments and why they're relevent to your insights, methods, etc.

The link above also references "bibtex" files. These are also extremely convenient, but have a steep learning curve and they make it difficult to tie them to an insight. If you use bibtext, then make sure that you provide a sentence to describe the source and it's relevance when you cite it - don't just add citations to the end of a sentence (this is common practice in academia, but I want to know that your citations are directly relevant for this assignmnet).
